A semantics of core AUTOSAR

### 2016-05-26 paper/paper.tex
Paper: TABs eliminated
Paper: Fixed rte command arities
PaJa+JoNo meeting

\documentclass[twocolumn]{article}
\usepackage{verbatim}
\usepackage[utf8x]{inputenc}
\usepackage[a4paper,margin=1.4cm]{geometry}
\usepackage[fleqn]{amsmath}
\usepackage{mathpartir}
% \usepackage{minted}
\title{A semantics of core AUTOSAR}
\author{Johan Nordlander \and Patrik Jansson}
%%%%%%%
\input{macros.tex}
\setlength{\parindent}{0pt}
\setlength{\parskip}{4pt plus 2pt minus 1pt}

@@

@@

The process terms of our calculus are as follows:
\begin{itemize}
\item $\Trunnable{\Tr{R}{I}}{K}{T}{\Act}{N}$  \newline
Common state for runnable $R$ of component $I$ with continuation $K$, showing $N$ current instances, $T$ seconds left until next possible instantiation, and activation state $\Act$.
\item $\Trinst{\Tr{R}{I}}{C}{\Xs}{\Code}$ \newline
An instance of runnable $R$ of component $I$, currently executing $\Code$ and owning the exclusive areas $\Xs$, possibly running on behalf of client $C$.
\item $\Texcl{\Tx{X}{I}}{B}$ \newline
Exclusive area $X$ of component $I$, with boolean occupation state $B$.
\item $\Tirv{\Ts{S}{I}}{V}$ \newline
Inter-runnable variable $S$ of component $I$, currently holding value $V$.
\item $\Tdelem{\Tep{E}{P}{I}}{B}{V}$ \newline
Non-queued sender-receiver data element $E$ in port $P$ of component $I$, currently holding value $V$ with boolean update status flag $B$.
\item $\Tqelem{\Tep{E}{P}{I}}{N}{\Vs}$ \newline
Queued sender-receiver data element $E$ in port $P$ of component $I$, currently holding value sequence $\Vs$. TODO Maximum queue capacity? $N$.
\item $\Topres{\Top{O}{P}{I}}{\Vs}$ \newline
Client-server operation $O$ in port $P$ of component $I$, currently holding result value sequence $\Vs$.
\item $\Ttimer{\Tr{R}{I}}{T}$ \newline  %TODO: timer has only 2 args in the semantics so {T_p} is not described
Countdown timer state for runnable $R$ of component $I$ with $T$ seconds left of its current cycle.
\end{itemize}

@@

The code part of a runnable instance should technically be a representation of the C or Matlab function that must accompany an AUTOSAR runnable definition once it is complete. However, since our task in this report is not to investigate the semantics of these languages in every detail, a more abstract notion of executable code will be required. To this end, we have chosen to ignore all details of internal computations, and just capture the observable effects of runnable execution as a pure sequence of RTE (VFB) calls. Because the result from an RTE call may in general affect subsequent runnable behavior, we use a continuation-passing style to allow any data-dependencies to be expressed.
%TODO: several LaTeX commands here are used with more arguments than in the semantics below
\[
\begin{array}{lll}
	\Code	& ::=	& \Enter{X}{\Cont}			\\
			& |	& \Exit{X}{\Cont}			\\
			& |	& \IrvWrite{S}{V}{\Cont}		\\
			& |	& \IrvRead{S}{\Cont}			\\
			& |	& \Send{E}{P}{V}{\Cont}		\\
			& |	& \Receive{E}{P}{\Cont}		\\
			& |	& \Write{E}{P}{V}{\Cont}		\\
			& |	& \Read{E}{P}{\Cont}		\\
			& |	& \Invalidate{E}{P}{\Cont}		\\
			& |	& \IsUpdated{E}{P}{\Cont}		\\
			& |	& \Call{O}{P}{V}{\Cont}		\\
			& |	& \Result{O}{P}{\Cont}		\\
			& |	& \Return{V}				\\
	\\
	\Cont	& \in	& V \rightarrow \Code		\\
\end{array}
\]

@@

@@

Two processes $P$ and $Q$ can make transitions in parallel if they agree on what is being said or heard. Just as the saying/hearing intuition suggests, at most one of the processes can have the saying role in such an agreement. The following set of inference rules capture this intuition formally.
\begin{mathpar}
	\inferrule
		{P \red{A!D} P' \\ Q \red{A?D} Q'}
		{P | Q \red{A!D} P' | Q'}
	\and
	\inferrule
		{P \red{A?D} P' \\ Q \red{A!D} Q'}
		{P | Q \red{A!D} P' | Q'}
	\and
	\inferrule
		{P \red{A?D} P' \\ Q \red{A?D} Q'}
		{P | Q \red{A?D} P' | Q'}
\end{mathpar}

@@

\item For each \emph{TimingEvent} triggering each runnable $R$ of each atomic component $I$, a term
\[
\begin{array}{c}
	\Ttimer{\Tr{R}{I}}{0.0}
\end{array}
\]
% where $T$ is the period of the event.
%TODO: Does this mean all timed runnables will want to act _immediately_ (at the same time)? No init delay?

@@

TODO: links from rules and choices made here to the AUTOSAR spec

@@

\onecolumn
\appendix
\section{Prolog code}
\label{sec:Prolog}

\input{semantics.tex}
{\renewcommand{\Prule}[2]{#1 & \Pif & #2\\}

\[ \begin{array}{r@{}c@{}lcr@{}c@{}lr@{}c@{}l}  \CombRed  \end{array} \]
}
\subsection{Exclusive areas}
\[ \begin{array}{r@{}c@{}l}      \ExclusiveAreas          \end{array} \]
\subsection{Inter-runnable variables}
\[ \begin{array}{r@{}c@{}l}      \InterrunnableVariables  \end{array} \]
\subsection{Sending/receiving}
\[ \begin{array}{r@{}c@{}l}      \SendingReceiving        \end{array} \]
%TODO: RuleLong?
\subsection{Reading/writing (unbuffered versions of rcv and snd)}
\[ \begin{array}{r@{}c@{}lcl}    \ReadingWriting          \end{array} \]
\subsection{Calling a server}
\[ \begin{array}{r@{}c@{}lcl}    \CallServer              \end{array} \]
\subsection{Obtaining a server result}
\[ \begin{array}{r@{}c@{}lcl}    \ServerResult            \end{array} \]
\subsection{Spawning and terminating}
\[ \begin{array}{r@{}c@{}lcl}    \SpawnTerminate          \end{array} \]
\subsection{Passing time}
\[ \begin{array}{r@{}c@{}lcl}    \PassingTime             \end{array} \]
\subsection{Ignoring broadcasts}
\[ \begin{array}{r@{}c@{}lcl}    \IgnoreBroadcast         \end{array} \]

>>>>

\documentclass[twocolumn]{article}
\usepackage{verbatim}
\usepackage[utf8x]{inputenc}
\usepackage[a4paper,margin=1.4cm]{geometry}
\usepackage[fleqn]{amsmath}
\usepackage{mathpartir}
\usepackage{ amssymb }
% \usepackage{minted}
\title{A semantics of core AUTOSAR}
\author{Johan Nordlander \and Patrik Jansson}
%%%%%%%
\input{macros.tex}
\setlength{\parindent}{0pt}
\setlength{\parskip}{4pt plus 2pt minus 1pt}

@@

%TODO: replace varibles with \V{Variable} (to make sure the style matches the semantic rules in the appendix - lower case variables, for example)

@@

The process terms of our calculus are as follows:
\begin{itemize}
\item $\Trunnable{\Tr{R}{I}}{K}{T}{\Act}{N}$  \newline
Common state for runnable $R$ of component $I$ with continuation $K$, showing $N$ current instances, $T$ seconds left until next possible instantiation, and activation state $\Act$.
\item $\Trinst{\Tr{R}{I}}{C}{\Xs}{\Code}$ \newline
An instance of runnable $R$ of component $I$, currently executing $\Code$ and owning the exclusive areas $\Xs$, possibly running on behalf of client $C$.
\item $\Texcl{\Tx{X}{I}}{B}$ \newline
Exclusive area $X$ of component $I$, with boolean occupation state $B$.
\item $\Tirv{\Ts{S}{I}}{V}$ \newline
Inter-runnable variable $S$ of component $I$, currently holding value $V$.
\item $\Tdelem{\Tep{E}{P}{I}}{B}{V}$ \newline
Non-queued sender-receiver data element $E$ in port $P$ of component $I$, currently holding value $V$ with boolean update status flag $B$.
\item $\Tqelem{\Tep{E}{P}{I}}{N}{\Vs}$ \newline
Queued sender-receiver data element $E$ in port $P$ of component $I$ with maximum capacity $N$, currently holding value sequence $\Vs$.
\item $\Topres{\Top{O}{P}{I}}{\Vs}$ \newline
Client-server operation $O$ in port $P$ of component $I$, currently holding result value sequence $\Vs$.
\item $\Ttimer{\Tr{R}{I}}{\V{T_p}}{T}$ \newline
Countdown timer with period \V{T_p} for runnable $R$ of component $I$ with $T$ seconds left of its current cycle.
\end{itemize}

@@

The code part of a runnable instance should technically be a representation of the C or Matlab function that must accompany an AUTOSAR runnable definition once it is complete. However, since our task in this report is not to investigate the semantics of these languages in every detail, a more abstract notion of executable code will be required. To this end, we have chosen to ignore all details of internal computations, and just capture the observable effects of runnable execution as a pure sequence of RTE (VFB) calls. Because the result from an RTE call may in general affect subsequent runnable behavior, we use a continuation-passing style to allow any data-dependencies to be expressed.
%
\[
\begin{array}{lll}
  \Code & ::= & \Enter{X}{\Cont}                \\
        & |   & \Exit{X}{\Cont}                 \\
        & |   & \IrvWrite{S}{V}{\Cont}          \\
        & |   & \IrvRead{S}{\Cont}              \\
        & |   & \Send{\Te{E}{P}}{V}{\Cont}      \\
        & |   & \Receive{\Te{E}{P}}{\Cont}      \\
        & |   & \Write{\Te{E}{P}}{V}{\Cont}     \\
        & |   & \Read{\Te{E}{P}}{\Cont}         \\
        & |   & \Invalidate{\Te{E}{P}}{\Cont}   \\
        & |   & \IsUpdated{\Te{E}{P}}{\Cont}    \\
        & |   & \Call{\To{O}{P}}{V}{\Cont}      \\
        & |   & \Result{\To{O}{P}}{\Cont}       \\
        & |   & \Return{V}                      \\
\\
  \Cont & \in & V \rightarrow \Code             \\
\end{array}
\]

@@

TODO: replace | with \Opar

@@

Two processes $P$ and $Q$ can make transitions in parallel if they agree on what is being said or heard. Just as the saying/hearing intuition suggests, at most one of the processes can have the saying role in such an agreement. The following set of inference rules capture this intuition formally.
\begin{mathpar}
  \inferrule
    {P \red{A!D} P' \\ Q \red{A?D} Q'}
    {\Tpar{P}{Q} \red{A!D} \Tpar{P'}{Q'}}
\and
  \inferrule
    {P \red{A?D} P' \\ Q \red{A!D} Q'}
    {P | Q \red{A!D} P' | Q'}
\and
  \inferrule
    {P \red{A?D} P' \\ Q \red{A?D} Q'}
    {P | Q \red{A?D} P' | Q'}
\end{mathpar}

@@

\item For each \emph{TimingEvent} triggering each runnable $R$ of each atomic component $I$, a term
\[
\begin{array}{c}
  \Ttimer{\Tr{R}{I}}{\V{T_p}}{0.0}
\end{array}
\]
where \V{T_p} is the period of the event.
%TODO: This means that all timed runnables want to act _immediately_ (at the same time - with no init delay).

@@

TODO: some links from rules and choices made here to the AUTOSAR spec

@@

\onecolumn
\appendix
\section{Semantic rules}
\label{sec:Prolog}

\input{semantics.tex}
{
\renewcommand{\Prule}[2]{#1 & \Pif & #2\\}
%\renewcommand{\Pcomma}{,\quad}
\renewcommand{\Pcomma}{\quad\text{and}\quad}

%TODO: fix the alignment of p2, q1
\[ \begin{array}{r@{}c@{}lcr@{}c@{}l@{}r@{}c@{}l}  \CombRed  \end{array} \]
}
\subsection{Exclusive areas}
\[ \begin{array}{r@{}c@{}l}      \ExclusiveAreas          \end{array} \]
\subsection{Inter-runnable variables}
\[ \begin{array}{r@{}c@{}l}      \InterrunnableVariables  \end{array} \]
\subsection{Sending/receiving}
\[ \begin{array}{r@{}c@{}l}      \SendingReceiving        \end{array} \]
%TODO: RuleLong?
\subsection{Reading/writing (unbuffered versions of rcv and snd)}
\[ \begin{array}{r@{}c@{}lcl}    \ReadingWriting          \end{array} \]
\subsection{Calling a server}
\[ \begin{array}{r@{}c@{}lcl}    \CallServer              \end{array} \]
\subsection{Obtaining a server result}
\[ \begin{array}{r@{}c@{}lcl}    \ServerResult            \end{array} \]
\subsection{Spawning and terminating}
%TODO: make the parallell composition more visible
\[ \begin{array}{r@{}c@{}lcl}    \SpawnTerminate          \end{array} \]
\subsection{Passing time}
\[ \begin{array}{r@{}c@{}lcl}    \PassingTime             \end{array} \]
\subsection{Ignoring broadcasts}
\[ \begin{array}{r@{}c@{}lcl}    \IgnoreBroadcast         \end{array} \]
